Tests for Measurement Invariance When Subgroups Are Unknown

### 2012-07-05 imps2012/invariance.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@145 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor changes to IMPS slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@146 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
added tests for complete ordering
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@147 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Added ordinal simulation results to imps slides.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@148 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{frame}[fragile]
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by the auxiliary variable.\\ \ \\
        \item Define $t \in (0, 1)$.
          The {\em empirical cumulative score process} is defined by:\\
  \begin{equation*}
    \bm{B}(\hat{\bm{\theta}}; t) = 
      {\bm{\widehat I}}^{-1/2} \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} s(\hat{\bm{\theta}} ; \bm{x}_i).
    \end{equation*}
  where $\lfloor nt \rfloor$ is the integer part of $nt$
  and $\bm{\widehat I}$ is some consistent covariance matrix estimate, e.g.,
  the observed information matrix ${\bm{I}}(\widehat{{\bm{\theta}}})$.

@@

\begin{frame}[fragile]
  \frametitle{Tests}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Under the hypothesis of
      measurement invariance, a functional central limit
      theorem holds:
  \begin{equation*}
  {\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
  \end{equation*}
 where ${\bm B}^{0}(\cdot)$ is a $p$-dimensional Brownian bridge.\\ \ \\

@@

@@

% Achim: Can you outline ordmax and possibly ordwmax here?
\frame{
  \frametitle{Test Statistics}
}

\frame{
  \frametitle{Simulation}
  \begin{itemize}
    \item Simulation:
      \begin{itemize}
        \item Same two-factor model from before.
        \item Measurement invariance violation in three unique
          variance parameters, growing as we move up the ordinal
          variable.
        \item Sample size in $\{120, 480, 960\}$.
        \item Levels of the ordinal variable in $\{4, 8, 12\}$.
        \item Five test statistics.
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Simulation}
  % Graph of results
}

% Wicherts example removed due to time constraints (about 15min)

@@

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Detailed examination of test properties.\\ \ \\
    \item Extension to related psychometric issues.\\ \ \\
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.
  \end{itemize}}

>>>>

\begin{frame}[fragile]
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by the auxiliary variable.\\ \ \\
        \item Define $t \in (0, 1)$.
          The {\em empirical cumulative score process} is defined by:\\
  \begin{equation*}
    \bm{B}(t; \hat{\bm{\theta}}) = 
      {\bm{\widehat I}}^{-1/2} \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} s(\hat{\bm{\theta}} ; \bm{x}_i).
    \end{equation*}
  where $\lfloor nt \rfloor$ is the integer part of $nt$
  and $\bm{\widehat I}$ is some consistent covariance matrix estimate, e.g.,
  the observed information matrix ${\bm{I}}(\widehat{{\bm{\theta}}})$.

@@

\begin{frame}[fragile]
  \frametitle{Tests}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Under the hypothesis of
      measurement invariance, a functional central limit
      theorem holds:
  \begin{equation*}
  {\bm{B}}(\cdot; \widehat{{\bm{\theta}}}) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
  \end{equation*}
 where ${\bm B}^{0}(\cdot)$ is a $p$-dimensional Brownian bridge.\\ \ \\

@@

\begin{frame}[fragile]
  \frametitle{Tests}

\begin{itemize}
  \item Test statistics aggregate over parameters $j = 1, \dots, k$ and observations $i = 1, \dots, n$,
    employing different norms.
  \item Special cases:
    \begin{eqnarray*}
      \mathit{DM}      & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
      \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
      \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
    \end{eqnarray*}
  \item ${\bm B}(\hat {\bm \theta})_{ij}$ is short for $\bm{B}(i/n; \hat{\bm{\theta}})_j$.
\end{itemize}

\end{frame}

@@

\frame{
  \frametitle{Simulation}
  \begin{itemize}
    \item Simulation:
      \begin{itemize}
        \item Same two-factor model from before.
        \item Measurement invariance violation in three unique
          variance parameters, growing as we move up the ordinal
          variable.
        \item Sample size in $\{120, 480, 960\}$.
        \item Levels of the ordinal variable in $\{4, 8, 12\}$.
        \item Three test statistics (two new statistics, plus the
          multiple group LRT).
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Simulation}
  % Graph of results
  \begin{center}
      \includegraphics[height=3in]{../ordinal/ordsim-xyplot.pdf}
  \end{center}
}

% library("lattice")
% load("simres2.rda")
% res$nlevels <- factor(paste("nlev=",res$nlevels,sep=""),
%   levels=c("nlev=4","nlev=8","nlev=12"))
% res$nobs <- paste("n=",res$nobs,sep="")

% pdf("ordsim-xyplot.pdf")
% trellis.par.set(theme = canonical.theme(color = FALSE))
% mykey <- simpleKey(c("LRT", "ordmaxLM", "orddwmax"), points = FALSE, lines = TRUE)
% mykey$lines$lty <- c(2, 1, 3)
% mykey$lines$lwd <- c(1.2, 1.2, 2)
% print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data=
% res[res$test %in% c("lrt","ordmax","ordwmax"),], type="l", xlab="Violation Magnitude", ylab="Power",key=mykey, ylim=c(0,1), lty=c(3,2,1), lwd=c(1.5,1.3,1)))

% dev.off()

@@

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Detailed examination of test properties.\\ \ \\
    \item Extension to related psychometric issues.\\ \ \\
    %\item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Paper: Merkle, E.\ C.\ \& Zeileis, A.\ (accepted).  Tests of measurement invariance without subgroups:\ {A} generalization of classical methods.  {\em Psychometrika}.
  \end{itemize}}
